---
title: "03 mlr3 Models"
author: "Alina Apel"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_depth: 3
    toc_float: true
    code_folding: show
    number_sections: true
    theme: default
---

# Machine Learning Models

This script evaluates machine-learning models based on the sentence-embedding dimensions. The models are compared using the same fixed 25 × 4 repeated cross-validation structure as in the semantic kNN analysis.

The models include a featureless baseline model, Ridge Regression, Elastic Net, Support Vector Machine, an untuned Random Forest, and a tuned Random Forest. Because the outcome variable `Score` is binary, Ridge Regression and Elastic Net are used as regularized logistic classification models.

## Libraries

```{r ml-libraries, message=FALSE, warning=FALSE}
library(dplyr)
library(tidyr)
library(purrr)
library(tibble)
library(knitr)
library(kableExtra)
library(ggplot2)

library(mlr3)
library(mlr3verse)
library(glmnet)
library(mlr3learners)
library(mlr3pipelines)
library(mlr3tuning)
library(paradox)
```

## Settings

```{r ml-settings}
set.seed(2026)

# Runtime settings
inner_folds <- 5

n_evals_elastic_net <- 30
n_evals_svm <- 30
n_evals_rf_tuned <- 50

num_trees_rf <- 500

# Output folder
dir.create("results/ml", recursive = TRUE, showWarnings = FALSE)
```

## Load Data and Resampling Structure

```{r load-ml-data}
data_ml <- readRDS("data/data_ml.rds")

resampling_25x4 <- readRDS("resampling/resampling_25x4_mlr3.rds")

data_ml <- data_ml %>%
  mutate(
    Score = factor(as.character(Score), levels = c("0", "1"))
  )

dim_cols <- grep("^Dim_", names(data_ml), value = TRUE)

cat("Number of embedding features:", length(dim_cols))
```

## Classification Task

```{r create-ml-task}
task_ml <- TaskClassif$new(
  id = "vera_embeddings_ml",
  backend = data_ml %>%
    select(Score, all_of(dim_cols)),
  target = "Score",
  positive = "1"
)

task_ml
```

# Helper Functions

## Performance Metrics

```{r classification-metric-function}
calculate_classification_metrics <- function(results) {
  
  observed <- as.numeric(as.character(results$Score))
  predicted <- as.numeric(as.character(results$predicted_score))
  
  tp <- sum(observed == 1 & predicted == 1, na.rm = TRUE)
  tn <- sum(observed == 0 & predicted == 0, na.rm = TRUE)
  fp <- sum(observed == 0 & predicted == 1, na.rm = TRUE)
  fn <- sum(observed == 1 & predicted == 0, na.rm = TRUE)
  
  n <- tp + tn + fp + fn
  
  accuracy <- (tp + tn) / n
  
  sensitivity <- ifelse(
    (tp + fn) == 0,
    NA_real_,
    tp / (tp + fn)
  )
  
  specificity <- ifelse(
    (tn + fp) == 0,
    NA_real_,
    tn / (tn + fp)
  )
  
  precision <- ifelse(
    (tp + fp) == 0,
    NA_real_,
    tp / (tp + fp)
  )
  
  f1 <- ifelse(
    is.na(precision) | is.na(sensitivity) | (precision + sensitivity) == 0,
    NA_real_,
    2 * precision * sensitivity / (precision + sensitivity)
  )
  
  balanced_accuracy <- mean(c(sensitivity, specificity), na.rm = TRUE)
  
  observed_correct <- (tp + fn) / n
  observed_incorrect <- (tn + fp) / n
  predicted_correct <- (tp + fp) / n
  predicted_incorrect <- (tn + fn) / n
  
  expected_accuracy <- 
    observed_correct * predicted_correct +
    observed_incorrect * predicted_incorrect
  
  kappa <- ifelse(
    (1 - expected_accuracy) == 0,
    NA_real_,
    (accuracy - expected_accuracy) / (1 - expected_accuracy)
  )
  
  tibble(
    accuracy = accuracy,
    sensitivity = sensitivity,
    specificity = specificity,
    precision = precision,
    f1 = f1,
    balanced_accuracy = balanced_accuracy,
    kappa = kappa
  )
}
```

## Extract Metrics from mlr3 Resampling Results

```{r extract-mlr3-metrics}
extract_resample_metrics <- function(resample_result, model_name) {
  
  map_dfr(seq_along(resample_result$predictions()), function(i) {
    
    prediction_i <- resample_result$predictions()[[i]]
    
    tibble(
      Score = as.numeric(as.character(prediction_i$truth)),
      predicted_score = as.numeric(as.character(prediction_i$response))
    ) %>%
      calculate_classification_metrics() %>%
      mutate(
        model = model_name,
        resample_number = i
      )
  })
}
```

## Run or Load Resampling Results

```{r run-or-load-function}
run_or_load_resample <- function(
    result_file,
    time_file,
    learner,
    task,
    resampling,
    seed = 2026,
    store_models = FALSE
) {
  
  if (file.exists(result_file) && file.exists(time_file)) {
    
    resample_result <- readRDS(result_file)
    computation_time <- readRDS(time_file)
    
  } else {
    
    set.seed(seed)
    
    start_time <- Sys.time()
    
    resample_result <- resample(
      task = task,
      learner = learner,
      resampling = resampling,
      store_models = store_models
    )
    
    end_time <- Sys.time()
    
    computation_time <- end_time - start_time
    
    saveRDS(resample_result, result_file)
    saveRDS(computation_time, time_file)
  }
  
  print(computation_time)
  return(resample_result)
}
```

# Models

## Baseline Model

The baseline model predicts the majority class in the training data.

```{r baseline-model}
baseline_file <- "results/ml/res_baseline.rds"
baseline_time_file <- "results/ml/computation_time_baseline.rds"

lrn_baseline <- lrn(
  "classif.featureless",
  predict_type = "response"
)

lrn_baseline$id <- "Baseline"

res_baseline <- run_or_load_resample(
  result_file = baseline_file,
  time_file = baseline_time_file,
  learner = lrn_baseline,
  task = task_ml,
  resampling = resampling_25x4
)
```

## Ridge Regression

Ridge Regression is implemented as regularized logistic regression with `alpha = 0`. The regularization parameter `lambda` is selected internally by cross-validation within `cv_glmnet`.

```{r ridge-model}
ridge_file <- "results/ml/res_ridge.rds"
ridge_time_file <- "results/ml/computation_time_ridge.rds"

pipe_ridge <- 
  po("imputemean") %>>%
  po("scale") %>>%
  lrn(
    "classif.cv_glmnet",
    id = "glmnet_ridge",
    predict_type = "response",
    alpha = 0,
    nfolds = 5,
    type.measure = "class"
  )

lrn_ridge <- as_learner(pipe_ridge)
lrn_ridge$id <- "Ridge"

res_ridge <- run_or_load_resample(
  result_file = ridge_file,
  time_file = ridge_time_file,
  learner = lrn_ridge,
  task = task_ml,
  resampling = resampling_25x4
)
```

## Elastic Net

Elastic Net is implemented as regularized logistic regression. The regularization parameter `lambda` is selected internally by `cv_glmnet`, while the mixing parameter `alpha` is tuned using inner cross-validation.

```{r elastic-net-model}
elastic_net_file <- "results/ml/res_elastic_net.rds"
elastic_net_time_file <- "results/ml/computation_time_elastic_net.rds"

pipe_elastic_net <- 
  po("imputemean") %>>%
  po("scale") %>>%
  lrn(
    "classif.cv_glmnet",
    id = "glmnet_en",
    predict_type = "response",
    nfolds = 5,
    type.measure = "class"
  )

lrn_elastic_net_pipe <- as_learner(pipe_elastic_net)

search_space_elastic_net <- ps(
  glmnet_en.alpha = p_dbl(
    lower = 0.05,
    upper = 1.00
  )
)

lrn_elastic_net <- AutoTuner$new(
  learner = lrn_elastic_net_pipe,
  resampling = rsmp("cv", folds = inner_folds),
  measure = msr("classif.bacc"),
  search_space = search_space_elastic_net,
  terminator = trm("evals", n_evals = n_evals_elastic_net),
  tuner = tnr("random_search")
)

lrn_elastic_net$id <- "Elastic_Net"

res_elastic_net <- run_or_load_resample(
  result_file = elastic_net_file,
  time_file = elastic_net_time_file,
  learner = lrn_elastic_net,
  task = task_ml,
  resampling = resampling_25x4
)
```

## Support Vector Machine

The SVM uses a radial kernel. The hyperparameters `cost` and `gamma` are tuned using inner cross-validation.

```{r svm-model}
svm_file <- "results/ml/res_svm.rds"
svm_time_file <- "results/ml/computation_time_svm.rds"

pipe_svm <- 
  po("imputemean") %>>%
  po("scale") %>>%
  lrn(
    "classif.svm",
    id = "svm",
    predict_type = "response",
    type = "C-classification",
    kernel = "radial"
  )

lrn_svm_pipe <- as_learner(pipe_svm)

search_space_svm <- ps(
  svm.cost = p_dbl(
    lower = -3,
    upper = 3,
    trafo = function(x) 10^x
  ),
  svm.gamma = p_dbl(
    lower = -5,
    upper = 0,
    trafo = function(x) 10^x
  )
)

lrn_svm <- AutoTuner$new(
  learner = lrn_svm_pipe,
  resampling = rsmp("cv", folds = inner_folds),
  measure = msr("classif.bacc"),
  search_space = search_space_svm,
  terminator = trm("evals", n_evals = n_evals_svm),
  tuner = tnr("random_search")
)

lrn_svm$id <- "SVM_tuned"

res_svm <- run_or_load_resample(
  result_file = svm_file,
  time_file = svm_time_file,
  learner = lrn_svm,
  task = task_ml,
  resampling = resampling_25x4
)
```

## Untuned Random Forest

The untuned Random Forest uses fixed hyperparameters and 500 trees.

```{r random-forest-model}
rf_file <- "results/ml/res_rf_fixed.rds"
rf_time_file <- "results/ml/computation_time_rf_fixed.rds"

pipe_rf_fixed <- 
  po("imputemean") %>>%
  lrn(
    "classif.ranger",
    id = "rf_fixed",
    predict_type = "response",
    num.trees = num_trees_rf,
    importance = "permutation"
  )

lrn_rf_fixed <- as_learner(pipe_rf_fixed)
lrn_rf_fixed$id <- "RF_fixed"

res_rf_fixed <- run_or_load_resample(
  result_file = rf_file,
  time_file = rf_time_file,
  learner = lrn_rf_fixed,
  task = task_ml,
  resampling = resampling_25x4
)
```

## Tuned Random Forest

The tuned Random Forest uses inner cross-validation to tune `mtry` and `min.node.size`.

```{r tuned-random-forest-model}
rf_tuned_file <- "results/ml/res_rf_tuned.rds"
rf_tuned_time_file <- "results/ml/computation_time_rf_tuned.rds"

pipe_rf_tuned <- 
  po("imputemean") %>>%
  lrn(
    "classif.ranger",
    id = "rf_tuned",
    predict_type = "response",
    num.trees = num_trees_rf,
    importance = "permutation"
  )

lrn_rf_tuned_pipe <- as_learner(pipe_rf_tuned)

search_space_rf_tuned <- ps(
  rf_tuned.mtry = p_int(
    lower = 1,
    upper = length(task_ml$feature_names)
  ),
  rf_tuned.min.node.size = p_int(
    lower = 1,
    upper = 20
  )
)

lrn_rf_tuned <- AutoTuner$new(
  learner = lrn_rf_tuned_pipe,
  resampling = rsmp("cv", folds = inner_folds),
  measure = msr("classif.bacc"),
  search_space = search_space_rf_tuned,
  terminator = trm("evals", n_evals = n_evals_rf_tuned),
  tuner = tnr("random_search")
)

lrn_rf_tuned$id <- "RF_tuned"

res_rf_tuned <- run_or_load_resample(
  result_file = rf_tuned_file,
  time_file = rf_tuned_time_file,
  learner = lrn_rf_tuned,
  task = task_ml,
  resampling = resampling_25x4
)
```

# Results

## Combine Results

```{r combine-ml-results}
baseline_metrics <- extract_resample_metrics(
  resample_result = res_baseline,
  model_name = "Baseline"
)

ridge_metrics <- extract_resample_metrics(
  resample_result = res_ridge,
  model_name = "Ridge Regression"
)

elastic_net_metrics <- extract_resample_metrics(
  resample_result = res_elastic_net,
  model_name = "Elastic Net"
)

svm_metrics <- extract_resample_metrics(
  resample_result = res_svm,
  model_name = "Tuned SVM"
)

rf_metrics <- extract_resample_metrics(
  resample_result = res_rf_fixed,
  model_name = "Untuned Random Forest"
)

rf_tuned_metrics <- extract_resample_metrics(
  resample_result = res_rf_tuned,
  model_name = "Tuned Random Forest"
)

ml_metrics_all <- bind_rows(
  baseline_metrics,
  ridge_metrics,
  elastic_net_metrics,
  svm_metrics,
  rf_metrics,
  rf_tuned_metrics
)

saveRDS(ml_metrics_all, "results/ml/ml_metrics_all.rds")
```

## Computation Time Summary

```{r computation-time-summary}
read_computation_time <- function(file) {
  if (file.exists(file)) {
    round(as.numeric(readRDS(file), units = "mins"), 2)
  } else {
    NA_real_
  }
}

computation_time_summary <- tibble(
  model = c(
    "Baseline",
    "Ridge Regression",
    "Elastic Net",
    "Tuned SVM",
    "Untuned Random Forest",
    "Tuned Random Forest"
  ),
  time_file = c(
    baseline_time_file,
    ridge_time_file,
    elastic_net_time_file,
    svm_time_file,
    rf_time_file,
    rf_tuned_time_file
  )
) %>%
  mutate(
    computation_time_minutes = map_dbl(time_file, read_computation_time)
  ) %>%
  select(model, computation_time_minutes)

computation_time_summary %>%
  knitr::kable(
    digits = 2,
    caption = "Computation time of the machine-learning models.",
    col.names = c("Model", "Computation time in minutes")
  ) %>%
  kableExtra::kable_styling(
    full_width = FALSE,
    bootstrap_options = c("hover")
  )
```


# ROC Analysis

ROC AUC was calculated as a measure of class separation. For the semantic kNN model, the mean neighbor score was used as the continuous prediction score. For the tuned SVM, class probabilities were used.

```{r roc-final-models, message=FALSE, warning=FALSE, fig.width=8, fig.height=6}
library(pROC)

dir.create("results/roc", recursive = TRUE, showWarnings = FALSE)

if (!exists("knn_predictions_all")) {
  knn_predictions_all <- readRDS("results/knn/knn_predictions_all.rds")
}

knn_roc_data <- knn_predictions_all %>%
  mutate(
    k = as.numeric(k),
    Score = as.numeric(as.character(Score))
  ) %>%
  filter(k == 15) %>%
  transmute(
    model = "Semantic kNN (k = 15)",
    resample_number,
    Score,
    score = neighbor_score_mean
  )

svm_roc_file <- "results/roc/svm_roc_data.rds"

if (file.exists(svm_roc_file)) {
  
  svm_roc_data <- readRDS(svm_roc_file)
  
} else {
  
  pipe_svm_prob <- 
    po("imputemean") %>>%
    po("scale") %>>%
    lrn(
      "classif.svm",
      id = "svm",
      predict_type = "prob",
      type = "C-classification",
      kernel = "radial"
    )
  
  lrn_svm_prob <- AutoTuner$new(
    learner = as_learner(pipe_svm_prob),
    resampling = rsmp("cv", folds = inner_folds),
    measure = msr("classif.bacc"),
    search_space = ps(
      svm.cost = p_dbl(lower = -2, upper = 3, trafo = function(x) 10^x),
      svm.gamma = p_dbl(lower = -3, upper = 1, trafo = function(x) 10^x)
    ),
    terminator = trm("evals", n_evals = n_evals_svm),
    tuner = tnr("random_search")
  )
  
  set.seed(2026)
  
  res_svm_prob <- resample(
    task = task_ml,
    learner = lrn_svm_prob,
    resampling = resampling_25x4,
    store_models = FALSE
  )
  
  svm_roc_data <- map_dfr(seq_along(res_svm_prob$predictions()), function(i) {
    prediction_i <- res_svm_prob$predictions()[[i]]
    
    tibble(
      model = "Tuned SVM",
      resample_number = i,
      Score = as.numeric(as.character(prediction_i$truth)),
      score = prediction_i$prob[, "1"]
    )
  })
  
  saveRDS(svm_roc_data, svm_roc_file)
}

roc_data <- bind_rows(knn_roc_data, svm_roc_data) %>%
  mutate(
    model = factor(model, levels = c("Semantic kNN (k = 15)", "Tuned SVM"))
  )

roc_auc_summary <- roc_data %>%
  group_by(model, resample_number) %>%
  summarise(
    roc_auc = as.numeric(
      pROC::auc(
        response = Score,
        predictor = score,
        levels = c(0, 1),
        direction = "<",
        quiet = TRUE
      )
    ),
    .groups = "drop"
  )

roc_auc_summary %>%
  group_by(model) %>%
  summarise(
    M = mean(roc_auc, na.rm = TRUE),
    SD = sd(roc_auc, na.rm = TRUE),
    Min = min(roc_auc, na.rm = TRUE),
    Max = max(roc_auc, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  knitr::kable(
    digits = 3,
    caption = "ROC AUC of the final semantic kNN model and the final tuned SVM model.",
    col.names = c("Model", "M", "SD", "Min", "Max"),
    align = c("l", "c", "c", "c", "c")
  ) %>%
  kableExtra::kable_styling(
    full_width = FALSE,
    bootstrap_options = c("hover"),
    font_size = 16
  ) %>%
  kableExtra::row_spec(row = 0, bold = TRUE)

roc_curve_data <- roc_data %>%
  group_by(model) %>%
  group_modify(~ {
    roc_object <- pROC::roc(
      response = .x$Score,
      predictor = .x$score,
      levels = c(0, 1),
      direction = "<",
      quiet = TRUE
    )
    
    tibble(
      false_positive_rate = 1 - roc_object$specificities,
      sensitivity = roc_object$sensitivities
    )
  }) %>%
  ungroup()

ggplot(roc_curve_data, aes(x = false_positive_rate, y = sensitivity, color = model)) +
  geom_abline(
    intercept = 0,
    slope = 1,
    linetype = "dashed",
    color = "grey50"
  ) +
  geom_line(linewidth = 1.2) +
  scale_color_manual(
    name = NULL,
    values = c(
      "Semantic kNN (k = 15)" = "#5B9BD5",
      "Tuned SVM" = "#ED7D31"
    )
  ) +
  scale_x_continuous(limits = c(0, 1), breaks = seq(0, 1, by = 0.1)) +
  scale_y_continuous(limits = c(0, 1), breaks = seq(0, 1, by = 0.1)) +
  labs(
    title = "ROC Curves of the Final Models",
    subtitle = "Pooled predictions across 100 train-test splits",
    x = "False positive rate",
    y = "Sensitivity"
  ) +
  theme_minimal(base_size = 16) +
  theme(
    plot.title = element_text(face = "bold", size = 22),
    plot.subtitle = element_text(size = 15),
    axis.title = element_text(size = 18, face = "bold"),
    axis.text = element_text(size = 14),
    panel.border = element_rect(color = "grey30", fill = NA, linewidth = 0.8),
    panel.grid.minor = element_blank(),
    legend.position = c(0.65, 0.15),
    legend.background = element_rect(fill = "white", color = "grey70"),
    legend.text = element_text(size = 13)
  )
```



# Final Model Comparison

Based on the model comparison, the tuned Support Vector Machine was selected as the final machine-learning model. It was compared with the final semantic kNN model from the previous analysis, which used `k = 15`.

```{r final-model-prepare}
chosen_ml_model <- "Tuned SVM"
chosen_k <- 15

# Load ML results if needed
if (!exists("ml_metrics_all")) {
  ml_metrics_all <- readRDS("results/ml/ml_metrics_all.rds")
}

# Load kNN results if needed
if (!exists("knn_metrics_all")) {
  
  knn_paths <- c(
    "results/knn/knn_metrics_all.rds",
    "results/knn_metrics_all.rds",
    "knn_metrics_all.rds"
  )
  
  knn_file <- knn_paths[file.exists(knn_paths)][1]
  
  if (is.na(knn_file)) {
    stop("Could not find knn_metrics_all. Please save the kNN results first or check the file path.")
  }
  
  knn_metrics_all <- readRDS(knn_file)
}

# Check available model names
unique(ml_metrics_all$model)

# Final SVM results
svm_metrics_final <- ml_metrics_all %>%
  filter(model == chosen_ml_model) %>%
  mutate(model = "Tuned SVM") %>%
  select(
    model,
    resample_number,
    accuracy,
    sensitivity,
    specificity,
    precision,
    f1,
    balanced_accuracy,
    kappa
  )

# Final semantic kNN results
knn_metrics_final <- knn_metrics_all %>%
  mutate(k = as.numeric(k)) %>%
  filter(k == chosen_k) %>%
  mutate(model = "Semantic kNN (k = 15)") %>%
  select(
    model,
    resample_number,
    accuracy,
    sensitivity,
    specificity,
    precision,
    f1,
    balanced_accuracy,
    kappa
  )

# Combine only the two final models
final_model_metrics <- bind_rows(
  knn_metrics_final,
  svm_metrics_final
) %>%
  mutate(
    model = factor(
      model,
      levels = c("Semantic kNN (k = 15)", "Tuned SVM")
    )
  )

# Second human rater benchmark
second_rater_metrics <- data_ml %>%
  transmute(
    Score = as.numeric(as.character(Score)),
    predicted_score = as.numeric(as.character(Score_2))
  ) %>%
  calculate_classification_metrics()

majority_class_baseline <- max(prop.table(table(data_ml$Score)))
second_rater_accuracy <- second_rater_metrics$accuracy

final_metric_order <- c(
  "Accuracy",
  "Balanced accuracy",
  "Cohen's kappa",
  "F1 score",
  "Precision",
  "Sensitivity",
  "Specificity"
)
```

## Final Model Performance

```{r final-model-performance-table}
final_model_summary <- final_model_metrics %>%
  select(
    model,
    accuracy,
    balanced_accuracy,
    kappa,
    f1,
    precision,
    sensitivity,
    specificity
  ) %>%
  pivot_longer(
    cols = -model,
    names_to = "Metric",
    values_to = "Value"
  ) %>%
  mutate(
    Metric = recode(
      Metric,
      accuracy = "Accuracy",
      balanced_accuracy = "Balanced accuracy",
      kappa = "Cohen's kappa",
      f1 = "F1 score",
      precision = "Precision",
      sensitivity = "Sensitivity",
      specificity = "Specificity"
    ),
    Metric = factor(Metric, levels = final_metric_order),
    model = factor(
      model,
      levels = c("Semantic kNN (k = 15)", "Tuned SVM")
    )
  ) %>%
  group_by(Metric, model) %>%
  summarise(
    M = mean(Value, na.rm = TRUE),
    SD = sd(Value, na.rm = TRUE),
    Min = min(Value, na.rm = TRUE),
    Max = max(Value, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  arrange(Metric, model)

final_model_summary %>%
  knitr::kable(
    digits = 3,
    caption = "Classification performance of the final semantic kNN model and the final tuned SVM model.",
    col.names = c("Metric", "Model", "M", "SD", "Min", "Max"),
    align = c("l", "l", "c", "c", "c", "c")
  ) %>%
  kableExtra::kable_styling(
    full_width = TRUE,
    bootstrap_options = c("hover", "condensed"),
    font_size = 15
  ) %>%
  kableExtra::row_spec(
    row = 0,
    bold = TRUE,
    background = "#F2F2F2"
  ) %>%
  kableExtra::row_spec(
    row = seq(2, nrow(final_model_summary), by = 2),
    extra_css = "border-bottom: 2px solid #D9D9D9;"
  ) %>%
  kableExtra::column_spec(
    column = 1,
    bold = TRUE,
    width = "14em"
  ) %>%
  kableExtra::column_spec(
    column = 2,
    width = "18em"
  ) %>%
  kableExtra::collapse_rows(
    columns = 1,
    valign = "middle"
  )
```

## Second Human Rater Benchmark

```{r final-model-second-rater-table}
second_rater_metrics %>%
  pivot_longer(
    cols = everything(),
    names_to = "Metric",
    values_to = "Value"
  ) %>%
  mutate(
    Metric = recode(
      Metric,
      accuracy = "Accuracy",
      sensitivity = "Sensitivity",
      specificity = "Specificity",
      precision = "Precision",
      f1 = "F1 score",
      balanced_accuracy = "Balanced accuracy",
      kappa = "Cohen's kappa"
    ),
    Metric = factor(Metric, levels = final_metric_order)
  ) %>%
  arrange(Metric) %>%
  knitr::kable(
    digits = 3,
    caption = "Second human rater performance compared with the original human score.",
    col.names = c("Metric", "Value"),
    align = c("l", "c")
  ) %>%
  kableExtra::kable_styling(
    full_width = FALSE,
    bootstrap_options = c("hover"),
    font_size = 16
  ) %>%
  kableExtra::row_spec(
    row = 0,
    bold = TRUE
  )
```


## Performance of the Final Tuned SVM

```{r final-svm-performance-plot, fig.width=10, fig.height=6}
svm_plot_data <- ml_metrics_all %>%
  filter(model == "Tuned SVM") %>%
  select(
    resample_number,
    accuracy,
    sensitivity,
    specificity,
    precision,
    f1,
    balanced_accuracy,
    kappa
  ) %>%
  pivot_longer(
    cols = -resample_number,
    names_to = "metric",
    values_to = "value"
  ) %>%
  mutate(
    metric = factor(
      metric,
      levels = c(
        "accuracy",
        "sensitivity",
        "specificity",
        "precision",
        "f1",
        "balanced_accuracy",
        "kappa"
      ),
      labels = c(
        "Accuracy",
        "Sensitivity",
        "Specificity",
        "Precision",
        "F1 score",
        "Balanced accuracy",
        "Cohen’s kappa"
      )
    )
  )

svm_mean_data <- svm_plot_data %>%
  group_by(metric) %>%
  summarise(
    value = mean(value, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  mutate(type = "Mean")

second_rater_plot_data <- tibble(
  metric = factor(
    c(
      "Accuracy",
      "Sensitivity",
      "Specificity",
      "Precision",
      "F1 score",
      "Balanced accuracy",
      "Cohen’s kappa"
    ),
    levels = levels(svm_plot_data$metric)
  ),
  value = c(0.959, 0.957, 0.961, 0.950, 0.954, 0.959, 0.917),
  type = "Second Human Rater"
)

legend_levels <- c("Single Trial", "Mean", "Second Human Rater")

ggplot(svm_plot_data, aes(x = metric, y = value, group = metric)) +
  geom_boxplot(
    width = 0.42,
    fill = "white",
    color = "grey30",
    outlier.shape = NA
  ) +
  geom_jitter(
    aes(
      color = factor("Single Trial", levels = legend_levels),
      shape = factor("Single Trial", levels = legend_levels),
      fill = factor("Single Trial", levels = legend_levels)
    ),
    width = 0.12,
    height = 0,
    alpha = 0.35,
    size = 2
  ) +
  geom_point(
    data = svm_mean_data,
    aes(
      x = metric,
      y = value,
      color = factor(type, levels = legend_levels),
      shape = factor(type, levels = legend_levels),
      fill = factor(type, levels = legend_levels)
    ),
    inherit.aes = FALSE,
    size = 4
  ) +
  geom_point(
    data = second_rater_plot_data,
    aes(
      x = metric,
      y = value,
      color = factor(type, levels = legend_levels),
      shape = factor(type, levels = legend_levels),
      fill = factor(type, levels = legend_levels)
    ),
    inherit.aes = FALSE,
    size = 3
  ) +
  scale_color_manual(
    name = NULL,
    breaks = legend_levels,
    values = c(
      "Single Trial" = "#5B9BD5",
      "Mean" = "black",
      "Second Human Rater" = "firebrick"
    )
  ) +
  scale_shape_manual(
    name = NULL,
    breaks = legend_levels,
    values = c(
      "Single Trial" = 16,
      "Mean" = 23,
      "Second Human Rater" = 16
    )
  ) +
  scale_fill_manual(
    name = NULL,
    breaks = legend_levels,
    values = c(
      "Single Trial" = "#5B9BD5",
      "Mean" = "gold",
      "Second Human Rater" = "firebrick"
    )
  ) +
  guides(
    color = guide_legend(
      override.aes = list(
        shape = c(16, 23, 16),
        fill = c("#5B9BD5", "gold", "firebrick"),
        alpha = c(0.35, 1, 1),
        size = c(2, 4, 3)
      )
    ),
    shape = "none",
    fill = "none"
  ) +
  scale_y_continuous(
    limits = c(0, 1.05),
    breaks = seq(0, 1, by = 0.1)
  ) +
  labs(
    title = "Final Tuned SVM Performance",
    subtitle = "Across 100 train-test splits",
    x = NULL,
    y = "Metric value"
  ) +
  theme_minimal(base_size = 16) +
  theme(
    plot.title = element_text(face = "bold", size = 22),
    plot.subtitle = element_text(size = 15),
    axis.title = element_text(size = 18, face = "bold"),
    axis.text.x = element_text(size = 14, angle = 30, hjust = 1),
    axis.text.y = element_text(size = 14),
    panel.grid.major.x = element_blank(),
    panel.grid.minor.x = element_blank(),
    panel.grid.minor.y = element_line(linewidth = 0.25),
    panel.border = element_rect(color = "grey30", fill = NA, linewidth = 0.8),
    legend.position = c(0.73, 0.15),
    legend.background = element_rect(fill = "white", color = "grey70"),
    legend.key = element_blank(),
    legend.text = element_text(size = 13)
  )

## Accuracy of the Final Models

```{r final-model-accuracy-plot, fig.width=9, fig.height=6}
accuracy_plot_data <- final_model_metrics %>%
  select(model, resample_number, accuracy) %>%
  mutate(
    x_pos = ifelse(model == "Semantic kNN (k = 15)", 1, 1.6)
  )

mean_accuracy_data <- accuracy_plot_data %>%
  group_by(model, x_pos) %>%
  summarise(
    accuracy = mean(accuracy, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  mutate(point_type = "Mean")

legend_levels <- c("Single Trial", "Mean")

ggplot(accuracy_plot_data, aes(x = x_pos, y = accuracy, group = model)) +
  geom_hline(
    yintercept = 0.50,
    linetype = "dashed",
    linewidth = 0.7,
    color = "grey40"
  ) +
  geom_hline(
    yintercept = majority_class_baseline,
    linetype = "dotted",
    linewidth = 0.9,
    color = "grey40"
  ) +
  geom_hline(
    yintercept = 0.90,
    linetype = "dashed",
    linewidth = 0.8,
    color = "grey40"
  ) +
  geom_hline(
    yintercept = second_rater_accuracy,
    linetype = "solid",
    linewidth = 1,
    color = "firebrick"
  ) +
  geom_boxplot(
    width = 0.35,
    fill = "white",
    color = "grey30",
    outlier.shape = NA
  ) +
  geom_jitter(
    aes(
      color = factor("Single Trial", levels = legend_levels),
      shape = factor("Single Trial", levels = legend_levels),
      fill = factor("Single Trial", levels = legend_levels)
    ),
    width = 0.08,
    height = 0,
    alpha = 0.35,
    size = 2
  ) +
  geom_point(
    data = mean_accuracy_data,
    aes(
      x = x_pos,
      y = accuracy,
      color = factor(point_type, levels = legend_levels),
      shape = factor(point_type, levels = legend_levels),
      fill = factor(point_type, levels = legend_levels)
    ),
    inherit.aes = FALSE,
    size = 4
  ) +
  annotate(
    "text",
    x = 2.15,
    y = second_rater_accuracy,
    label = "Second human rater",
    hjust = 1,
    vjust = -0.4,
    size = 4.5,
    color = "firebrick"
  ) +
  annotate(
    "text",
    x = 2.15,
    y = 0.90,
    label = "90% accuracy",
    hjust = 1,
    vjust = -0.4,
    size = 4.5
  ) +
  annotate(
    "text",
    x = 2.15,
    y = majority_class_baseline,
    label = "Majority-class baseline",
    hjust = 1,
    vjust = -0.4,
    size = 4.5
  ) +
  annotate(
    "text",
    x = 2.15,
    y = 0.50,
    label = "50% accuracy",
    hjust = 1,
    vjust = -0.4,
    size = 4.5
  ) +
  scale_color_manual(
    name = NULL,
    breaks = legend_levels,
    values = c(
      "Single Trial" = "#5B9BD5",
      "Mean" = "black"
    )
  ) +
  scale_shape_manual(
    name = NULL,
    breaks = legend_levels,
    values = c(
      "Single Trial" = 16,
      "Mean" = 23
    )
  ) +
  scale_fill_manual(
    name = NULL,
    breaks = legend_levels,
    values = c(
      "Single Trial" = "#5B9BD5",
      "Mean" = "gold"
    )
  ) +
  guides(
    color = guide_legend(
      override.aes = list(
        shape = c(16, 23),
        fill = c("#5B9BD5", "gold"),
        alpha = c(0.35, 1),
        size = c(2, 4)
      )
    ),
    shape = "none",
    fill = "none"
  ) +
  scale_x_continuous(
    breaks = c(1, 1.6),
    labels = c("Semantic kNN (k = 15)", "Tuned SVM"),
    limits = c(0.75, 2.2)
  ) +
  scale_y_continuous(
    limits = c(0.45, 1.02),
    breaks = seq(0.45, 1.00, by = 0.05)
  ) +
  coord_cartesian(clip = "off") +
  labs(
    title = "Accuracy of the Final Models",
    subtitle = "Semantic kNN with k = 15 and tuned SVM across 100 train-test splits",
    x = NULL,
    y = "Accuracy"
  ) +
  theme_minimal(base_size = 16) +
  theme(
    plot.title = element_text(face = "bold", size = 22),
    plot.subtitle = element_text(size = 15),
    axis.title = element_text(size = 18, face = "bold"),
    axis.text.x = element_text(size = 14),
    axis.text.y = element_text(size = 14),
    panel.grid.major.x = element_blank(),
    panel.grid.minor.x = element_blank(),
    panel.grid.minor.y = element_line(linewidth = 0.25),
    panel.border = element_rect(color = "grey30", fill = NA, linewidth = 0.8),
    legend.position = c(0.05, 0.07),
    legend.justification = c(0, 0),
    legend.background = element_rect(fill = "white", color = "grey70"),
    legend.key = element_blank(),
    legend.text = element_text(size = 13),
    plot.margin = margin(5.5, 45, 5.5, 5.5)
  )
```

